---
title: "BIOSC1203 Geomorphometric Analysis"
format:
  html:
    page-layout: full
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: sandstone
    include-in-header:
      text: |
        <style>
        /* ── Tab strip spacing ───────────────────────────────────────── */
        .nav-tabs {
          gap: 4px;
          flex-wrap: wrap;
          border-bottom: 2px solid #c8a96e;
        }

        .nav-tabs .nav-item {
          margin-bottom: -2px;
        }

        /* Default (inactive) tab */
        .nav-tabs .nav-link {
          background-color: #f5f0e8;
          color: #5a4a2f;
          border: 1px solid #d4c5a0;
          border-bottom: none;
          border-radius: 6px 6px 0 0;
          padding: 6px 14px;
          font-size: 0.85rem;
          font-weight: 500;
          transition: background-color 0.15s, color 0.15s;
          white-space: nowrap;
        }

        .nav-tabs .nav-link:hover {
          background-color: #e8dfc8;
          color: #3b2f1a;
          border-color: #c8a96e;
        }

        /* Active (selected) tab */
        .nav-tabs .nav-link.active {
          background-color: #8c6d3f;
          color: #ffffff;
          border-color: #8c6d3f;
          border-bottom-color: #8c6d3f;
          font-weight: 600;
        }

        /* Tab content panel */
        .tab-content {
          border: 1px solid #d4c5a0;
          border-top: none;
          border-radius: 0 0 6px 6px;
          padding: 16px;
          background: #ffffff;
        }

        /* ── GPA run spinner ─────────────────────────────────────────── */
        #gpa-running {
          display: none;
          align-items: center;
          gap: 12px;
          margin-top: 10px;
          padding: 10px 14px;
          background: #f5f0e8;
          border: 1px solid #c8a96e;
          border-radius: 6px;
          font-size: 0.85rem;
          color: #5a4a2f;
        }

        .gpa-spinner {
          width: 20px;
          height: 20px;
          border: 3px solid #d4c5a0;
          border-top-color: #8c6d3f;
          border-radius: 50%;
          animation: spin 0.8s linear infinite;
          flex-shrink: 0;
        }

        @keyframes spin {
          to { transform: rotate(360deg); }
        }
        </style>
        <script>
        $(document).on('click', '#run_tps_analysis', function() {
          $('#gpa-running').css('display', 'flex');
        });
        $(document).on('shiny:idle', function() {
          $('#gpa-running').css('display', 'none');
        });
        </script>
server: shiny
---

```{r setup, include=FALSE}
required_pkgs <- c("geomorph", "shiny", "DT", "ggplot2", "plotly", "readr", "dplyr", "stringr")
new_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
if (length(new_pkgs) > 0) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

library(geomorph)
library(shiny)
library(DT)
library(ggplot2)
library(plotly)
library(readr)
library(dplyr)
library(stringr)

source("geomorph_helpers.R")

# Path where student metadata is saved between sessions.
# The file lives next to about.qmd on the server.
METADATA_FILE <- "student_metadata.csv"
```

## Overview

This tool has five sections:

1. **Upload & Preview** — load your 3D Slicer CSV, classify landmarks, preview
2. **Download TPS** — convert to TPS format and download
3. **Ecological Metadata** — record habitat, activity period, and locomotion for your specimen; saved so you can reload it later
4. **TPS Analysis** — upload one or more TPS files for multi-specimen GPA + PCA
5. **Geomorph Analyses** — results from the TPS analysis

---

## 1 · Upload & Preview

```{r ui-upload}
#| panel: sidebar

fileInput("csv_file", "Upload CSV file",
  accept = c(".csv", "text/csv"),
  buttonLabel = "Browse...",
  placeholder = "No file selected"
)
hr()

textInput("specimen_name", "Specimen name", value = "Genus_species",
  placeholder = "e.g. Homo_sapiens")
helpText("Use Genus_species format. This becomes the ID= field in the TPS file
          and the key used to save/load your metadata.")

hr()

h5("Column mapping")
helpText("Match your CSV column names to the fields below.")
uiOutput("col_pickers")

hr()

actionButton("run_parse", "Parse landmarks", class = "btn-primary", width = "100%")
```

```{r ui-preview}
#| panel: fill

# Raw data tab always visible once CSV is uploaded
conditionalPanel(
  condition = "output.tbl_raw_ready",
  tabsetPanel(
    tabPanel("Raw data",                DTOutput("tbl_raw")),
    tabPanel("Parsed landmarks",        DTOutput("tbl_parsed")),
    tabPanel("Landmark classification", verbatimTextOutput("lm_summary"))
  )
)
```

---

## 2 · Download TPS

```{r ui-tps}
#| panel: sidebar

helpText("Once landmarks are parsed, generate the TPS file.")
hr()
downloadButton("dl_tps", "Download TPS file", class = "btn-success")
```

```{r ui-tps-preview}
#| panel: fill

h5("TPS file preview")
verbatimTextOutput("tps_preview")
```

---

## 3 · Ecological Metadata

```{r ui-meta}
#| panel: sidebar

h5("Load saved metadata")
helpText("Enter your specimen name above (Section 1), then click Load to
          pull back anything you saved in a previous session.")
actionButton("load_meta", "Load saved metadata", width = "100%")

hr()

h5("Enter metadata for this specimen")

textAreaInput("meta_habitat",
  label    = "Habitat",
  rows     = 3,
  placeholder = "Describe where this species lives — e.g. tropical rainforest
understory, rocky arid scrubland, open savanna, freshwater riparian..."
)

textAreaInput("meta_activity",
  label    = "Activity period",
  rows     = 2,
  placeholder = "When is this species active? — e.g. diurnal, nocturnal,
crepuscular, cathemeral..."
)

textAreaInput("meta_locomotion",
  label    = "Locomotion / movement through environment",
  rows     = 3,
  placeholder = "How does this species move? — e.g. arboreal climbing,
suspensory, terrestrial quadruped, gliding, swimming, saltatorial..."
)

textAreaInput("meta_notes",
  label    = "Additional notes",
  rows     = 3,
  placeholder = "Anything else relevant — diet, body size, phylogenetic
notes, observer name..."
)

hr()

actionButton("save_meta", "Save metadata", class = "btn-success", width = "100%")
uiOutput("meta_save_status")
```

```{r ui-meta-out}
#| panel: fill

h5("Saved metadata for all specimens")
helpText("All entries saved in this project session are shown below.
          The table is written to student_metadata.csv next to about.qmd
          so it persists across Quarto preview restarts.")
DTOutput("tbl_meta")
hr()
downloadButton("dl_meta", "Download metadata CSV", class = "btn-outline-secondary")
```

---

## 4 · TPS File Upload & Analysis

```{r ui-tps-upload}
#| panel: sidebar

h5("Upload TPS file(s)")
helpText("Upload a single TPS file that contains one or more specimens
          (each block: LM3=n / coordinates / ID=name).
          You can combine multiple student TPS files into one before uploading.")

fileInput("tps_file", "Upload TPS file",
  accept      = c(".tps", "text/plain"),
  buttonLabel = "Browse...",
  placeholder = "No file selected"
)

hr()

h5("GPA settings")
checkboxInput("tps_has_sliders", "Include semilandmarks as sliders?", value = TRUE)
selectInput("tps_procD", "Procrustes distance method",
  choices  = c("full" = "full", "orthogonal" = "orthogonal"),
  selected = "full"
)

hr()

h5("PCA settings")
numericInput("tps_n_pcs", "Number of PCs to display", value = 3, min = 2, max = 10)

hr()

actionButton("run_tps_analysis", "Run GPA + PCA", class = "btn-primary", width = "100%")

# Spinner shown while analysis runs, hidden when Shiny goes idle
HTML('<div id="gpa-running"><div class="gpa-spinner"></div>Running GPA + PCA — this may take a moment...</div>')
```

```{r ui-tps-analysis-out}
#| panel: fill

# Specimens in TPS always shown once file is uploaded
conditionalPanel(
  condition = "output.tbl_tps_ready",
  tabsetPanel(
    tabPanel("Specimens in TPS",   DTOutput("tbl_tps_specimens")),
    tabPanel("GPA summary",        verbatimTextOutput("tps_gpa_summary")),
    tabPanel("PCA scores",         DTOutput("tbl_tps_pca")),
    tabPanel("PCA plot",           plotlyOutput("plot_tps_pca",    height = "500px")),
    tabPanel("Variance explained", plotlyOutput("plot_tps_scree",  height = "400px")),
    tabPanel("Consensus shape",    plotOutput("plot_tps_consensus", height = "500px"))
  )
)
```

---

```{r server}
#| context: server

# ── helper functions ──────────────────────────────────────────────────────────

read_csv_flexible <- function(path) {
  tryCatch({
    df <- read_csv(path, show_col_types = FALSE)
    if (ncol(df) < 3) df <- read_delim(path, delim = ";", show_col_types = FALSE)
    df
  }, error = function(e) NULL)
}

classify_landmarks <- function(df, label_col, x_col, y_col, z_col) {
  df |>
    select(label = all_of(label_col),
           x     = all_of(x_col),
           y     = all_of(y_col),
           z     = all_of(z_col)) |>
    mutate(
      label = as.character(label),
      label = if_else(is.na(label) | label == "",
                      paste0("unlabeled_", row_number()), label),
      x = as.numeric(x),
      y = as.numeric(y),
      z = as.numeric(z),
      type = case_when(
        str_detect(label, "^OC_")        ~ "semi",
        str_detect(label, "^unlabeled_") ~ "semi",
        TRUE                              ~ "fixed"
      )
    ) |>
    filter(!is.na(x), !is.na(y), !is.na(z))
}

# TPS format:  LM3=n / x  y  z (tab-delimited, width-padded) / ID=name
build_tps <- function(lm_df, specimen_name) {
  n   <- nrow(lm_df)
  fmt <- function(v) formatC(v, format = "f", digits = 4, width = 12, flag = " ")
  coord_lines <- paste(fmt(lm_df$x), fmt(lm_df$y), fmt(lm_df$z), sep = "  ")
  c(paste0("LM3=", n), coord_lines, paste0("ID=", specimen_name))
}

# ── metadata helpers ──────────────────────────────────────────────────────────

# Load the metadata table from disk (or return empty tibble if not yet created)
load_metadata_table <- function() {
  if (file.exists(METADATA_FILE)) {
    tryCatch(
      read_csv(METADATA_FILE, show_col_types = FALSE),
      error = function(e) tibble(
        specimen = character(), habitat = character(),
        activity = character(), locomotion = character(), notes = character(),
        saved_at = character()
      )
    )
  } else {
    tibble(
      specimen   = character(),
      habitat    = character(),
      activity   = character(),
      locomotion = character(),
      notes      = character(),
      saved_at   = character()
    )
  }
}

# Upsert one row (replace existing entry for this specimen, or append)
upsert_metadata <- function(tbl, specimen, habitat, activity, locomotion, notes) {
  new_row <- tibble(
    specimen   = specimen,
    habitat    = habitat,
    activity   = activity,
    locomotion = locomotion,
    notes      = notes,
    saved_at   = format(Sys.time(), "%Y-%m-%d %H:%M:%S")
  )
  tbl <- tbl |> filter(specimen != !!specimen)
  bind_rows(tbl, new_row)
}

# ── section 1 & 2: CSV upload / parse / TPS ───────────────────────────────────

raw_data <- reactive({
  req(input$csv_file)
  read_csv_flexible(input$csv_file$datapath)
})

output$col_pickers <- renderUI({
  df <- raw_data()
  req(df)
  cols <- names(df)
  guess <- function(patterns) {
    m <- cols[str_detect(tolower(cols), paste(tolower(patterns), collapse = "|"))]
    if (length(m)) m[1] else cols[1]
  }
  tagList(
    selectInput("col_label", "Label column", choices = cols,
                selected = guess(c("label","name","id","lm"))),
    selectInput("col_x",     "X / R coord",  choices = cols,
                selected = guess(c("^r$","_x$","x$","r$"))),
    selectInput("col_y",     "Y / A coord",  choices = cols,
                selected = guess(c("^a$","_y$","y$","a$"))),
    selectInput("col_z",     "Z / S coord",  choices = cols,
                selected = guess(c("^s$","_z$","z$","s$")))
  )
})

output$tbl_raw <- renderDT({
  req(raw_data())
  datatable(raw_data(), options = list(scrollX = TRUE, pageLength = 10),
            rownames = FALSE)
})

# Flag used by conditionalPanel to show tabs only after CSV is uploaded
output$tbl_raw_ready <- reactive({ !is.null(raw_data()) })
outputOptions(output, "tbl_raw_ready", suspendWhenHidden = FALSE)

parsed <- eventReactive(input$run_parse, {
  req(raw_data(), input$col_label, input$col_x, input$col_y, input$col_z)
  classify_landmarks(raw_data(), input$col_label,
                     input$col_x, input$col_y, input$col_z)
})

output$tbl_parsed <- renderDT({
  req(parsed())
  df <- parsed() |> mutate(across(c(x, y, z), \(v) round(v, 4)))
  datatable(df, options = list(scrollX = TRUE, pageLength = 15),
            rownames = FALSE) |>
    formatStyle("type",
      backgroundColor = styleEqual(
        c("fixed", "semi"), c("#d4edda", "#cce5ff")))
})

output$lm_summary <- renderPrint({
  req(parsed())
  df <- parsed()
  cat("Total landmarks:", nrow(df), "\n")
  cat("Fixed landmarks:", sum(df$type == "fixed"), "\n")
  cat("Semilandmarks  :", sum(df$type == "semi"),  "\n\n")
  cat("Fixed landmark labels:\n")
  cat(paste(" ", df$label[df$type == "fixed"], collapse = "\n"), "\n\n")
  cat("Semilandmark labels:\n")
  cat(paste(" ", df$label[df$type == "semi"],  collapse = "\n"), "\n")
})

tps_lines <- reactive({
  req(parsed())
  build_tps(parsed(), input$specimen_name)
})

output$tps_preview <- renderText({
  req(tps_lines())
  paste(tps_lines(), collapse = "\n")
})

output$dl_tps <- downloadHandler(
  filename = function() paste0(input$specimen_name, ".tps"),
  content  = function(file) writeLines(tps_lines(), file)
)

# ── section 3: ecological metadata ───────────────────────────────────────────

# Reactive value holding the in-memory metadata table (synced to disk on save)
meta_table <- reactiveVal(load_metadata_table())

# Load button: pull the saved row for this specimen into the text boxes
observeEvent(input$load_meta, {
  req(input$specimen_name)
  tbl <- meta_table()
  row <- tbl |> filter(specimen == input$specimen_name)
  if (nrow(row) == 1) {
    updateTextAreaInput(session, "meta_habitat",    value = row$habitat)
    updateTextAreaInput(session, "meta_activity",   value = row$activity)
    updateTextAreaInput(session, "meta_locomotion", value = row$locomotion)
    updateTextAreaInput(session, "meta_notes",      value = row$notes)
  } else {
    showNotification(
      paste0("No saved metadata found for '", input$specimen_name, "'."),
      type = "warning", duration = 4
    )
  }
})

# Save button: upsert into the table and write to disk
observeEvent(input$save_meta, {
  req(input$specimen_name)
  updated <- upsert_metadata(
    tbl        = meta_table(),
    specimen   = input$specimen_name,
    habitat    = input$meta_habitat,
    activity   = input$meta_activity,
    locomotion = input$meta_locomotion,
    notes      = input$meta_notes
  )
  meta_table(updated)
  tryCatch({
    write_csv(updated, METADATA_FILE)
    showNotification(
      paste0("Metadata saved for '", input$specimen_name, "'."),
      type = "message", duration = 3
    )
  }, error = function(e) {
    showNotification(
      paste0("Could not write to disk: ", e$message,
             ". Metadata is saved for this session only."),
      type = "warning", duration = 6
    )
  })
})

output$tbl_meta <- renderDT({
  tbl <- meta_table()
  validate(need(nrow(tbl) > 0, "No metadata saved yet."))
  datatable(tbl, options = list(scrollX = TRUE, pageLength = 10),
            rownames = FALSE)
})

output$dl_meta <- downloadHandler(
  filename = function() paste0("metadata_", Sys.Date(), ".csv"),
  content  = function(file) write_csv(meta_table(), file)
)

# ── section 4 & 5: TPS upload → GPA + PCA ────────────────────────────────────

# Parse an uploaded TPS file into a named list of coordinate matrices
# Returns list(coords_array = p×3×n array, specimen_ids = character vector,
#              lm_types = character vector of length p)
parse_tps_file <- function(path) {
  lines <- readLines(path, warn = FALSE)
  lines <- lines[nzchar(trimws(lines))]  # drop blank lines

  specimens  <- list()
  spec_ids   <- character()
  i          <- 1

  while (i <= length(lines)) {
    line <- trimws(lines[i])

    if (grepl("^LM3?=", line, ignore.case = TRUE)) {
      n_lm <- as.integer(sub("^LM3?=", "", line, ignore.case = TRUE))
      coords <- matrix(NA_real_, nrow = n_lm, ncol = 3)

      for (j in seq_len(n_lm)) {
        i <- i + 1
        vals <- as.numeric(strsplit(trimws(lines[i]), "\\s+")[[1]])
        coords[j, ] <- vals[1:3]
      }

      # Scan ahead for ID= line (skip any SCALE= lines)
      spec_id <- paste0("specimen_", length(specimens) + 1)
      while (i < length(lines)) {
        i <- i + 1
        next_line <- trimws(lines[i])
        if (grepl("^ID=", next_line, ignore.case = TRUE)) {
          spec_id <- sub("^ID=", "", next_line, ignore.case = TRUE)
          break
        }
        if (grepl("^LM3?=", next_line, ignore.case = TRUE)) {
          i <- i - 1   # back up — next iteration will catch this LM= line
          break
        }
      }

      specimens[[length(specimens) + 1]] <- coords
      spec_ids <- c(spec_ids, spec_id)

    } else {
      i <- i + 1
    }
  }

  if (length(specimens) == 0) return(NULL)

  p <- nrow(specimens[[1]])
  n <- length(specimens)

  # Build p × 3 × n array
  A <- array(NA_real_, dim = c(p, 3, n))
  for (k in seq_len(n)) A[, , k] <- specimens[[k]]
  dimnames(A)[[3]] <- spec_ids

  # Classify landmark type from first specimen's row count:
  # we don't have label info in a bare TPS, so type = "fixed" for all
  # (sliders are handled by index position passed to gpagen)
  list(A = A, ids = spec_ids, n_lm = p, n_spec = n)
}

tps_parsed <- eventReactive(input$tps_file, {
  req(input$tps_file)
  result <- tryCatch(
    parse_tps_file(input$tps_file$datapath),
    error = function(e) NULL
  )
  validate(need(!is.null(result), "Could not parse the TPS file. Check the format."))
  result
})

output$tbl_tps_specimens <- renderDT({
  req(tps_parsed())
  tp <- tps_parsed()
  df <- data.frame(
    Index    = seq_along(tp$ids),
    Specimen = tp$ids,
    N_landmarks = tp$n_lm
  )
  datatable(df, options = list(pageLength = 20), rownames = FALSE)
})

# Flag used by conditionalPanel to show tabs only after TPS is uploaded
output$tbl_tps_ready <- reactive({ !is.null(input$tps_file) })
outputOptions(output, "tbl_tps_ready", suspendWhenHidden = FALSE)

tps_results <- eventReactive(input$run_tps_analysis, {
  req(tps_parsed())
  tp <- tps_parsed()

  validate(
    need(tp$n_spec >= 2,
         paste0("GPA requires at least 2 specimens. Your TPS file has ",
                tp$n_spec, ". Combine more student TPS files into one and re-upload."))
  )

  # Build sliders for semilandmarks if requested.
  # Without label info in a raw TPS, we can't auto-detect semi vs fixed.
  # Sliders must be defined externally or disabled here.
  sliders <- NULL
  # (sliders will be wired up once A provides the row indices)

  gpa <- tryCatch(
    gpagen(tp$A, curves = sliders,
           ProcD = (input$tps_procD == "full"),
           print.progress = FALSE),
    error = function(e) {
      validate(need(FALSE, paste("GPA failed:", e$message)))
    }
  )

  pca <- tryCatch(
    gm.prcomp(gpa$coords),
    error = function(e) {
      validate(need(FALSE, paste("PCA failed:", e$message)))
    }
  )

  list(gpa = gpa, pca = pca, ids = tp$ids)
})

output$tps_gpa_summary <- renderPrint({
  validate(need(input$run_tps_analysis > 0, "Click 'Run GPA + PCA' to see results."))
  req(tps_results())
  res <- tps_results()
  cat("=== GPA (Generalized Procrustes Analysis) ===\n\n")
  print(summary(res$gpa))
  cat("\nConsensus shape (first 6 landmarks):\n")
  print(round(head(res$gpa$consensus, 6), 4))
})

output$tbl_tps_pca <- renderDT({
  validate(need(input$run_tps_analysis > 0, "Click 'Run GPA + PCA' to see results."))
  req(tps_results())
  pca     <- tps_results()$pca
  eigs    <- pca$sdev^2
  var_exp <- round(eigs / sum(eigs) * 100, 2)
  n_pcs   <- min(input$tps_n_pcs, length(var_exp))

  scores_df <- as.data.frame(pca$x[, 1:n_pcs, drop = FALSE])
  names(scores_df) <- paste0("PC", 1:n_pcs, " (", var_exp[1:n_pcs], "%)")
  scores_df$Specimen <- rownames(scores_df)
  scores_df <- scores_df[, c("Specimen", setdiff(names(scores_df), "Specimen"))]
  datatable(scores_df, options = list(scrollX = TRUE), rownames = FALSE)
})

output$plot_tps_pca <- renderPlotly({
  validate(need(input$run_tps_analysis > 0, "Click 'Run GPA + PCA' to see results."))
  req(tps_results())
  pca     <- tps_results()$pca
  eigs    <- pca$sdev^2
  var_exp <- round(eigs / sum(eigs) * 100, 2)
  scores  <- as.data.frame(pca$x)
  scores$Specimen <- rownames(scores)

  p <- ggplot(scores, aes(x = Comp1, y = Comp2, label = Specimen)) +
    geom_point(size = 3, colour = "#2c7bb6") +
    geom_text(vjust = -0.8, size = 3) +
    labs(
      x     = paste0("PC1 (", var_exp[1], "% variance)"),
      y     = paste0("PC2 (", var_exp[2], "% variance)"),
      title = "PCA morphospace — PC1 vs PC2"
    ) +
    theme_minimal(base_size = 13)

  ggplotly(p, tooltip = c("label", "x", "y"))
})

output$plot_tps_scree <- renderPlotly({
  validate(need(input$run_tps_analysis > 0, "Click 'Run GPA + PCA' to see results."))
  req(tps_results())
  pca     <- tps_results()$pca
  eigs    <- pca$sdev^2
  var_exp <- round(eigs / sum(eigs) * 100, 2)
  n_show  <- min(input$tps_n_pcs, length(var_exp))
  df_sc   <- data.frame(
    PC       = factor(paste0("PC", 1:n_show), levels = paste0("PC", 1:n_show)),
    Variance = var_exp[1:n_show]
  )

  p <- ggplot(df_sc, aes(x = PC, y = Variance)) +
    geom_col(fill = "#2c7bb6", alpha = 0.8) +
    geom_line(aes(group = 1), colour = "#d7191c", linewidth = 0.8) +
    geom_point(colour = "#d7191c", size = 2) +
    labs(title = "Scree plot — variance explained per PC",
         y = "% Variance", x = "") +
    theme_minimal(base_size = 13)

  ggplotly(p)
})

output$plot_tps_consensus <- renderPlot({
  validate(need(input$run_tps_analysis > 0, "Click 'Run GPA + PCA' to see results."))
  req(tps_results())
  cons    <- tps_results()$gpa$consensus
  df_plot <- data.frame(
    x     = cons[, 1],
    y     = cons[, 2],
    z     = cons[, 3],
    label = rownames(cons)
  )

  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

  plot(df_plot$x, df_plot$y, pch = 19, col = "#2c7bb6", cex = 1.2,
       xlab = "X (R)", ylab = "Y (A)", main = "Consensus — XY plane (R-A view)")
  text(df_plot$x, df_plot$y, labels = df_plot$label, cex = 0.6, pos = 3)

  plot(df_plot$x, df_plot$z, pch = 19, col = "#2c7bb6", cex = 1.2,
       xlab = "X (R)", ylab = "Z (S)", main = "Consensus — XZ plane (R-S view)")
  text(df_plot$x, df_plot$z, labels = df_plot$label, cex = 0.6, pos = 3)
})
```